---
title: "Linear Models"
format:
  revealjs:
    scrollable: true
    theme: [default, styles.scss]
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      src: notes/chalkboard_1a.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: false
    eval: true
    message: false
    warnings: false
    comment: "#>" 
    
revealjs-plugins:
  - pointer
  - verticator
  
filters: 
  - reveal-header
  - code-fullscreen
  - reveal-auto-agenda
  
auto-agenda:
  bullets: numbered
  heading: Learning Outcomes
  
editor: visual
---

```{r}
#| include: false

library(palmerpenguins)
library(ggformula)
library(ThemePark)
library(tidyverse)
library(taylor)
library(DT)
library(tidytuesdayR)
set.seed(29)
penguins_nona <- penguins |> drop_na() |> 
  filter(species != "Chinstrap") |> 
  slice_sample(n=8) |>
  select(species, bill_length_mm)
horrors <- tt_load(2022, week =44)$horror_movies
scooby <- tt_load(2021, week =29)$scoobydoo
```

# Motivation: Danceability

## Motivation: Danceability

```{r}
set.seed(66745675)
taylors <- taylor_all_songs |>
  filter(album_name %in% c("Midnights", "Fearless (Taylor's Version)")) |>
  select(album_name, track_name, danceability) |>
  drop_na() |> slice_sample(n=6) 
taylors |>  datatable(options = list(dom = "t"))
```

## Find the Empty Model

## Find the Empty Model

$$
\hat Y_i = 0.5698
$$

## Find SSE

$$
SSE = 0.0573
$$

## Find a Model explaining variability with album

## Find a Model explaining variability with album

$$
\hat Y_i = 0.519 + 0.1017 X
$$

$$
X = \left\{\begin{array}{cc}
1 & Midnights\\
0 & Other
\end{array}
\right.
$$

## Find the SSE of the previous Linear Model

## Find the SSE of the previous Linear Model

$$
SSE  = 0.0418
$$

# Motivation: Horror Movies

## Motivation: Horror Movies

```{r}
set.seed(4414)
horror <- horrors |> filter(genre_names %in% 
                    c("Horror, Thriller", "Horror, Mystery", "Horror, Horror")) |> 
  group_by(genre_names) |> 
  select(genre_names, title, vote_average) |>
  slice_sample(n=3) |> ungroup()
horror |> datatable(options = list(dom = "t"))
```

## Create the Empty Model for `vote_average`

## Create the Empty Model for `vote_average`

$$
\hat Y_i = 4.856
$$

## Find SSE for Empty Model

## Find SSE for Empty Model

$$
SSE = 36.562
$$

## Find a Model to explain `vote_average` by `genre_names`

## Find a Model to explain `vote_average` by `genre_names`

$$
\hat Y_i = 5.4667 -1.1 X - 0.733 Z  
$$

$$
X = \left\{\begin{array}{cc}
1 & Mystery \\
0 & Other
\end{array}
\right.
\ \ 
Z = \left\{\begin{array}{cc}
1 & Thriller\\
0 & Other
\end{array}
\right.
$$

## Find SSE of previous Linear Model

## Find SSE of previous Linear Model

$$
SSE = 34.680
$$

# Motivation: Scooby Doo

## Motivation: Scooby Doo

```{r}
set.seed(531334111)
doo <- scooby |> mutate(zoinks_count = as.numeric(zoinks)) |> drop_na() |>
  filter(zoinks_count>1) |> 
  filter(monster_type %in% 
           c("Possessed Object", "Mythical", "Undead", "Magician")) |>
  group_by(monster_type) |> 
  select(monster_type, zoinks_count) |> 
  slice_sample(n=2) |> ungroup()
doo |> datatable(options = list(dom = "t"))
```

## Fit an Empty Model on Zoinks Count

## Fit an Empty Model on Zoinks Count

$$
\hat Y_i = 3.375
$$

## SSE of Empty Model

## SSE of Empty Model

$$
SSE = 39.875
$$

## Fit a model explaining Zoinks count and monster type

## Fit a model explaining Zoinks count and monster type

$$
\hat Y_i = 2.5 -0.5 X +3 Z +1 A   
$$

$$
X = \left\{\begin{array}{cc}
1 & Mythical \\
0 & Other
\end{array}
\right.
\ \ 
Z = \left\{\begin{array}{cc}
1 & Possessed \ Object\\
0 & Other
\end{array}
\right.
\ \
A = \left\{\begin{array}{cc}
1 & Undead\\
0 & Other
\end{array}
\right.
$$

## Find the SSE of the previous model

## Find the SSE of the previous model

$$
SSE = 25.5
$$
