---
title: "Confidence Intervals"
format:
  revealjs:
    scrollable: true
    theme: [default, styles.scss]
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false
    code-fold: true
    chalkboard:
      src: notes/chalkboard_1a.json
      storage: chalkboard_pres
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: false
    eval: true
    message: false
    warnings: false
    comment: "#>" 
    error: true
    
revealjs-plugins:
  - pointer
  - verticator

filters: 
  - reveal-header
  - reveal-auto-agenda
  - webr
  
auto-agenda:
  bullets: numbered
  heading: Learning Outcomes
  
editor: visual
---

```{r}
#| include: false
#| cache: true

set.seed(3249)
res <- lm(body_mass_kg ~ species, data = penguins)
themes <- list(theme_bw(),
               theme(axis.title = element_text(size = 32),
        axis.text = element_text(size = 26)))
b1_sim_1000_1 <- tibble(b1 = rnorm(1000, 1, sd = 1)) + 3.7
b1_sim_1000_2 <- tibble(b1 = rnorm(1000, -0.1, sd = 1)) + 3.7
b1_sim_1000_3 <- tibble(b1 = rnorm(1000, -2, sd = 1)) + 3.7
b1_sim_1000_4 <- tibble(b1 = rnorm(1000, 3, sd = 1)) + 3.7
b1_boot_1000 <- tibble(b1 = rnorm(1000, 3.7, sd = 1))

```

# Confidence Intervals

## Motivating Example

```{r}
library(coursekata)
penguins |> ggplot(aes(x=species, y = body_mass_kg)) +
  geom_jitter() +
  theme(axis.title = element_text(size = 36)) +
  labs(x = "Species", y = "Body Mass") + themes
```

## Motivating Example

```{r}
penguins |> ggplot(aes(x=species, y = body_mass_kg)) +
  geom_jitter() + geom_boxplot() +
  theme(axis.title = element_text(size = 36)) +
  labs(x = "Species", y = "Body Mass") + themes
```

## Linear Models

```{r}
#| echo: true

res <- penguins %>% lm(body_mass_kg ~ species, data = .)
res
```

## Computing the mean

$$
\hat Y = 3.71 + 0.03 \times Chinstrap + 1.39 \times Gentoo 
$$

-   Adelie

-   Chinstrap

-   Gentoo

## Confidence Interval

These are a collection of $\beta_1$ values that will fail to reject the Null Hypothesis.

::: fragment
The $\beta_1$ values that are not different from our current estimate $b_1$
:::

## Confidence Interval

These are a collection of specific parameters that will fail to reject the null hypothesis.

-   Adelie: $\mu_a$

-   Chinstrap: $\mu_c$

-   Gentoo: $\mu_g$

## Looking at Adelie

```{r}
b1_sim_1000_4 |> ggplot(aes(b1)) + 
  geom_density() +
  geom_vline(xintercept = 3.7) +
  geom_density(data = b1_sim_1000_1) +
  geom_density(data = b1_sim_1000_2) +
  geom_density(data = b1_sim_1000_3) +
  geom_density(data = b1_boot_1000, 
               fill = "green", alpha = .2)  +
  xlab("Adelie") +
  themes
```

## Boostrap CI

```{r}
#| echo: true
extract_mean <- function(x){
means <- favstats(body_mass_kg ~ species, data = penguins) %>% 
  select(species, mean) %>%
  t() %>% .[2,] |> as.numeric() 
return(tibble(adelie = means[1], 
                chinstrap = means[2], 
                gentoo = means[3]))
}
boot_means <- do(1000) * extract_mean(resample(penguins))
print("Adelie")
quantile(boot_means$adelie, probs = c(.025, .975))
print("Chinstrap")
quantile(boot_means$chinstrap, probs = c(.025, .975))
print("Gentoo")
quantile(boot_means$gentoo, probs = c(.025, .975))
```

## Mathematical Models

Use the `pairwise()` to compute confidence intervals.

A difference occurs when 0 is not within the intervals.

```{r}
#| echo: true
res %>% pairwise()
```

## Plots

```{r}
#| echo: true

res %>% pairwise(plot = T)

```

# Semester

## Final

-   Max of 30 Questions

-   Mixed of Multiple Choice, True or False, and Free Response

-   Get 1 sheet of notes, front and back

-   In-person

-   You are not allowed to talk to anyone during the exam

-   You can use an inexpensive calculator, bring one

-   All other electronic devices must be silenced an put away

-   The exam is worth 100 points

-   Write any explanations in the context of the problem

## Final Material

-   Linear Models

-   Inference

-   Confidence Intervals

## MW 10 AM Class Final

Wednesday December 6

8 AM - 10 AM

## MW 2 PM Class Final

Wednesday December 6

1 PM - 3 PM

## When are Labs DUE

DEC 7th 11:59 PM

## When are Readings DUE

DEC 7th 11:59 PM

## When are Extra Credits DUE

DEC 7th 11:59 PM

# Course Evaluations

## Please Complete Course Evaluations

-   Complete them for all your courses

# Complete Survey

## Complete Survey

![](img/frame%20(7).png){fig-align="center"}
