---
title: "Logistic Regression"
date: 3/11/25
format:
  revealjs:
    width: 1200
    scrollable: true
    theme: [default, styles.scss]
    navigation-mode: vertical
    controls-layout: bottom-right
    controls-tutorial: true
    incremental: false 
    chalkboard:
      theme: whiteboard
      chalk-width: 4
knitr:
  opts_chunk: 
    echo: true
    eval: true
    message: false
    code-fold: true
    warnings: false
    comment: "#>" 
    
revealjs-plugins:
  - pointer
  - verticator
  
filters: 
  - reveal-header
  - code-fullscreen
  - reveal-auto-agenda
editor: source
---

```{r}
#| include: false

library(tidyverse)
library(csucistats)
library(MASS)
Melanoma$dead <- ifelse(Melanoma$status == 1, 1, 0)

theme_set(theme_bw() + 
            theme(
              axis.text.x = element_text(size = 24),
              axis.title = element_text(size = 30),
              plot.title = element_text(size = 48),
              strip.text = element_text(size = 20),
              legend.title = element_blank(),
              legend.text = element_text(size = 24)))

```

# Motivation

## Melanoma

Melanoma is a type of skin cancer that causes the cells that produce melanin to grow out of control. What makes Melanoma so dangerous is that it can metastasize to other parts of the body.

## Outcome of Interest

We are interested in learning how do different factors affect and individual's chances of survival. Therefore, we are measuring patients and if they lived or died during a study period.

## Data 

We will be using the `Melanoma` data set with the following variables: `dead` (died by Melanoma, 1=yes, 0= no),  `sex` (1 = male, 0 = female), `age` (in years), `thickness` (tumour thickness in mm), ulcer (1 = presence, 0 absent).

## Plot

```{r}
Melanoma |> 
  ggplot(aes(thickness, dead)) +
  geom_point()
```

# Modeling Binary Outcomes

## Construct Model

$$
\left(\begin{array}{c}
Dead \\
Alive
\end{array}\right) = \beta_0 + \beta_1X
$$

## Let ...

$$
Y = \left\{\begin{array}{cc}
1 & Dead \\
0 & Alive
\end{array}\right.
$$

## Construct a Model

$$
P\left(Y = 1\right) = \beta_0 + \beta_1X
$$

## Construct a Model

$$
P\left(Y = 1\right) = \frac{e^{\beta_0 + \beta_1X}}{1 + e^{\beta_0 + \beta_1X}}
$$

## Construct a Model

$$
\frac{P(Y = 1)}{P(Y = 0)} = e^{\beta_0 + \beta_1X}
$$

where $\frac{P(Y = 1)}{P(Y = 0)}$ are considered the odds of observing $Y = 1$.

## The Logistic Model

$$
\log\left\{\frac{P(Y = 1)}{P(Y = 0)}\right\} = \beta_0 + \beta_1X
$$


# Logistic Regression

## Logistic Regression

Logistic Regression is used to model the association between a set of predictors and a **binary** outcome variable.

::: fragment
This is similar Linear Regression which models the association between a set of predictors and a **numerical** outcome variable.
:::

## Logistic Regression

Logistic Regression uses the logistic model to formulate the relationship between the predictors and the outcome.

::: fragment

More specifically, for an outcome of Y:

$$
Y = \left\{\begin{array}{cc}
1 & \text{Category 1} \\
0 & \text{Category 2}
\end{array}\right.
$$

The Predictors variable will model the probability of observing category 1 ($P(Y=1)$)

:::


## Logistic Model

$$
\log\left\{\frac{P(Y = 1)}{P(Y = 0)}\right\} = \beta_0 + \beta_1X
$$

## Regression Coefficients $\beta$

The regression coefficients quantify how a specific predictor changes the odds of observing the first category of the outcome ($Y = 1$)

## Estimating $\beta$

To obtain the numerical value for $\beta$, denoted as $\hat \beta$, we will be finding the values of $\hat \beta$ that maximizes the likelihood function:

$$
L(\boldsymbol \beta) = \prod_{i=1}^n \left(\frac{e^{\beta_0 + \beta_1X}}{1 + e^{\beta_0 + \beta_1X}}\right)^{Y_i}\left(\frac{1}{1 + e^{\beta_0 + \beta_1X}}\right)^{1-Y_i}
$$

The likelihood function can be thought as the probability of observing the entire data set. Therefore, we want to choose the values the $\beta_0$ and $\beta_1$ that will result in the highest probability of observing the data.


## Estimated Parameters

The values you obtain ($\hat \beta$) tell you the relationship between the a predictor variable and the log odds of observing the first category of the outcome $Y=1$.

::: fragment
Exponentiating the estimate ($e^{\hat \beta}$) will give you the relationship between a predictor variable and the odds of observing the first category of the outcome $Y=1$.
:::

## Interpreting $\hat \beta$

For a continuous predictor variable:

As X increases by 1 unit, the odds of observing the first category ($Y = 1$) increases by a factor of $e^{\hat\beta}$.

For a categorical predictor variable (first dummy variable):

The odds of observing the first category ($Y = 1$) in the indicated category ($D=1$) is $e^{\hat\beta}$ times higher/lower compared to the reference category ($D=0$).


## Logistic Regression in R

```{r}
#| echo: true
#| eval: false

glm(Y ~ X1 + X2 + ... + Xp,
    data = DATA,
    family = binomial())
```


## Example

Modelling `dead` by `sex`, `age`, `thickness`, and `ulcer`: 

```{r}
glm(dead ~ sex + age + thickness + ulcer,
    data = Melanoma,
    family = binomial())
```

$$
\log(odds) = -2.40 + 0.41 M + 0.004 age + 0.11 thick + 1.31ulc
$$


## Interpretation

$$
\log(odds) = -2.40 + 0.41 M + 0.004 age + 0.11 thick + 1.31ulc
$$

::: fragment
The odds of experiencing death in males is 1.5 times larger than the odds for females, adjusting for age, tumor thickness and the presence of an ulcer.
:::

::: fragment
As age increases by 1 year, the odds of experiencing death increases by a factor of 1.004, adjusting for sex, tumor thickness and presence of an ulcer.
:::

# Prediction with Models

## Working with probabilities

As you can see, working with odds may be unintuitive for the average person. It will be better to predict the probability and display those results to individuals.

## Predicting Probability


$$
\hat P\left(Y = 1\right) = \frac{e^{\hat\beta_0 + \hat\beta_1X_1 + \cdots  + \hat \beta_pX_p}}{1 + e^{\hat\beta_0 + \hat\beta_1X_1 + \cdots  + \hat \beta_pX_p}}
$$

## Prediction in R

```{r}
#| echo: true
#| eval: false
xglm <- glm(Y ~ X1 + X2 + ... + Xp,
            data = DATA,
            family = binomial())
predict_df <- data.frame(X1 = VAL,  X2 = VAL, ..., Xp = VAL)
predict(xglm,
        predict_df,
        type = "response")
```


## Example 1

Predict the probability of observing death for a patient who is male, 75 years old, an tumor thickness of 2.9, and **with** ulcer presence. 

```{r}
xglm <- glm(dead ~ sex + age + thickness + ulcer,
    data = Melanoma,
    family = binomial())
predict_df <- data.frame(sex = 1, age = 75, thickness = 2.9, ulcer = 1)
predict(xglm, predict_df, type = "response")
```


## Example 1

Predict the probability of observing death for a patient who is male, 75 years old, an tumor thickness of 2.9, and **without** ulcer presence. 

```{r}
xglm <- glm(dead ~ sex + age + thickness + ulcer,
    data = Melanoma,
    family = binomial())
predict_df <- data.frame(sex = 1, age = 75, thickness = 2.9, ulcer = 0)
predict(xglm, predict_df, type = "response")
```


## Example 1

| | Ulcer Present | Ulcer not Present |
|:-|:-|:-|
| Probabiltiy | 48.7% | 20.3% |

For a patient who is 75 years old, male, and a tumor thickness of 2.9, when an ulcer is present, the probability a patient dies increases to 48% compared to a patient who does not have an ulcer with a probability of 20%.


## Example 2

Predict the probability of observing death for a patient who is female, **75** years old, an tumor thickness of 2.9, and with ulcer presence. 

```{r}
xglm <- glm(dead ~ sex + age + thickness + ulcer,
    data = Melanoma,
    family = binomial())
predict_df <- data.frame(sex = 0, age = 75, thickness = 2.9, ulcer = 1)
predict(xglm, predict_df, type = "response")
```


## Example 2

Predict the probability of observing death for a patient who is female, **55** years old, an tumor thickness of 2.9, and with ulcer presence. 

```{r}
xglm <- glm(dead ~ sex + age + thickness + ulcer,
    data = Melanoma,
    family = binomial())
predict_df <- data.frame(sex = 0, age = 55, thickness = 2.9, ulcer = 1)
predict(xglm, predict_df, type = "response")
```

## Example 2

| | 55 | 75 |
|:-|:-|:-|
| Probabiltiy | 36.8% | 38.7% |

For a patient who is female, a tumor thickness of 2.9, and has an ulcer, the probability a 75 year old patient dies increases to 38.7% compared to a 55 year old patient with a probability of 36.8%.

